### 2014-11-07 documents/paper/ms_v0.0.tex
current doc

\documentclass[12pt,preprint]{aastex}
\usepackage{amssymb,amsmath,mathrsfs,hyperref,datetime}

\newcommand{\datavector}[1]{\boldsymbol{#1}}
\newcommand{\data}{\datavector{y}}
\newcommand{\datum}{\data_i}
\newcommand{\truedatum}{\data_{{\rm true}, i}}
\newcommand{\noiselessdata}{\tilde{\data}}
\newcommand{\noiselessdatum}{\tilde{\datum}}
\newcommand{\xdcov}{\datavector{V}_j}
\newcommand{\datacov}{\datavector{C}}
\newcommand{\postcov}{\tilde{\datacov}_{i, j}}
\newcommand{\vmu}{\datavector{\mu}}
\newcommand{\postmu}{\tilde{\vmu}_{i, j}}
\newcommand{\datumcov}{\datacov_i}
\newcommand{\transpose}{{\rm T}}
\newcommand\rf[1]{{\bf [RF: #1]}}

\def\urltilda{\kern -.15em\lower .7ex\hbox{\~{}}\kern .04em}

\pdfoutput=1

\begin{document}

\title{Joint photometric modeling for improved measurements in imaging surveys.}
\author{Ross~Fadely\altaffilmark{1} \&
        David~W.~Hogg\altaffilmark{1,2}}
\altaffiltext{1}{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics, New~York~University, 4~Washington~Place, New~York, NY 10003, USA}
\altaffiltext{2}{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, 69117 Heidelberg, Germany}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Abstract
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Large scale astronomical imaging surveys have fundamentally changed our 
understanding of the cosmos, and promise to continue to do so with forthcoming 
surveys like the Large Synoptic Survey Telescope which will provide greater 
volumes of data at increasingly fainter detection limits.  In order to extract
the most science out of such costly missions, it is important to continually
explore ways to improve the means by which photometric measurements are made.
Historically, photometric measurements are made on a per object basis under
our understanding of the characteristics of the telescope, instruments, and
astronomical sources.  We argue that it is possible to improve the quality of
such photometric measurements by building models which incorporate the shared
similarity of sources.  As a demonstration, we use Extreme Deconvolution (XD)
to model the joint distribution of photometric quantities measured by SDSS in
Stripe 82.  Using single epoch data, we show that our joint XD model makes
predictions that more closely resemble the coadded Stripe 82 data.  Looking
at the difference between the psf and model magnitudes in the single epoch and 
coadded data, the average error is XX at $r\sim21$ versus an average of XX
for our XD posteriors.  In the case of the CMDs of globular clusters, we show
that the XD posteriors lessen the presence of misclassified galaxies and
present a narrower stellar locus.  We suggest a number of improvements that
can be made to provide improved photometric inferences.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Section - Introduction
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

Large imaging surveys of parts (or the full) sky are one main modes in which
modern astronomers study the universe.  Whether conducted on the ground (e.g.,
SDSS, UKIDSS, NEWFIRM, VLA, etc.) or in space (GALEX, WISE, XMM?, etc.), the
volume of such surveys provides both strong statistical samples of objects and
also the potential for new discoveries.  For instance, the Sload Digital Sky
Survey (optical $ugriz$, $r\lesssim 22$) \citep{york00} was first dataset able
to describe the quasar luminosity function (CITE?) while also enabling the 
discovery of new streams and satellites around the Milky Way (CITE).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Section - Method
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Extreme Deconvolution}

\begin{itemize}
\item Mixture of Gaussians where each datum's observational uncertainty is accounted for.
\item cite bovy and applicable works.
\item regularization to avoid singularities
\end{itemize}

\subsection{Posterior estimation}

After inference, the XD mixture provides an estimation for the noiseless distribution of 
the data:

\begin{eqnarray}\displaystyle
p(\noiselessdata) = \sum_j^K \frac{\alpha_j}{\sqrt{2\pi\det{|\xdcov|}}} \exp \left (-\frac{1}{2}(\noiselessdata - \vmu_j)^\transpose \xdcov^{-1}(\noiselessdata - \vmu_j) \right )
\quad .
\label{eqn:noiseless}
\end{eqnarray}

In a hierarchical sense, this learned distribution provides a prior distribution on what the 
value of a datum should be --- the distribution of the denoised data $\noiselessdata$ is 
learned for all the observations via XD.  Using Bayes' Rule, we can estimate the posterior 
value for a given datum

\begin{eqnarray}\displaystyle
p(\truedatum | \datum) & \propto & p(\datum | \truedatum) p(\truedatum) \\
                                         & \propto & \frac{1}{\sqrt{2\pi\det{|\datumcov|}}} \exp \left (-\frac{1}{2}(\datum - \truedatum)^\transpose \datumcov^{-1}(\datum - \truedatum) \right ) p(\truedatum)
\quad ,
\label{eqn:posterior}
\end{eqnarray}

\noindent where we assume that the prior for $\truedatum$ is $p(\noiselessdata)$.  Since 
$p(\noiselessdata)$ is a sum of Gaussian densities, and we multiply this by a Gaussian 
likelihood, the posterior is also a sum of Gaussians.  That is, the posterior for the true 
value of the observation is

\begin{eqnarray}\displaystyle
p(\truedatum | \datum) & \propto & \sum_j^K \frac{\alpha_j}{\sqrt{2\pi\det{|\postcov|}}} \exp \left (-\frac{1}{2}(\truedatum - \postmu)^\transpose \postcov^{-1}(\truedatum - \postmu) \right ) \\
{\rm where,}\\
\postcov & = & (\datumcov^{-1} + \xdcov^{-1})^{-1} \quad {\rm and,} \\
\postmu & = &  \postcov^{-1} (\datumcov^{-1} \datum + \xdcov^{-1} \vmu_j)
\label{eqn:posterior}
\end{eqnarray}
\end{document}

### 2014-11-07 code/utils.py
adding some code

@@ -1,0 +1,402 @@
+# 
+# Various routines, mostly for handing data.
+#
+# Author - Ross Fadely (unless otherwise noted).
+#
+import os
+import multiprocessing
+import numpy as np
+import pyfits as pf
+
+def log_multivariate_gaussian_Nthreads(x, mu, V, Nthreads=1):
+    """
+    Use multiprocessing to calculate log likelihoods.
+    """
+    n_samples = x.shape[0]
+
+    pool = multiprocessing.Pool(Nthreads)
+    mapfn = pool.map
+    Nchunk = np.ceil(1. / Nthreads * n_samples).astype(np.int)
+
+    if len(V.shape) > 3:
+        perdatum = True
+    else:
+        perdatum = False
+
+    arglist = [None] * Nthreads
+    for i in range(Nthreads):
+        s = i * Nchunk
+        e = s + Nchunk
+        if perdatum:
+            arglist[i] = (x[s:e], mu, V[s:e])
+        else:
+            arglist[i] = (x[s:e], mu, V)
+
+    result = list(mapfn(lmg, [args for args in arglist]))
+    
+    logls = result[0]
+    for i in range(1, Nthreads):
+       logls = np.vstack((logls, result[i]))
+       
+    pool.close()
+    pool.terminate()
+    pool.join()
+    return logls
+
+def lmg(args):
+    return log_multivariate_gaussian(*args)
+
+def log_multivariate_gaussian(x, mu, V, Vinv=None, method=1):
+    """
+    Swiped from astroML:
+    https://github.com/astroML/astroML/blob/master/astroML/utils.py
+
+    Evaluate a multivariate gaussian N(x|mu, V)
+
+    This allows for multiple evaluations at once, using array broadcasting
+
+    Parameters
+    ----------
+    x: array_like
+        points, shape[-1] = n_features
+
+    mu: array_like
+        centers, shape[-1] = n_features
+
+    V: array_like
+        covariances, shape[-2:] = (n_features, n_features)
+
+    Vinv: array_like or None
+        pre-computed inverses of V: should have the same shape as V
+
+    method: integer, optional
+        method = 0: use cholesky decompositions of V
+        method = 1: use explicit inverse of V
+
+    Returns
+    -------
+    values: ndarray
+        shape = broadcast(x.shape[:-1], mu.shape[:-1], V.shape[:-2])
+
+    Examples
+    --------
+
[…]

### 2014-11-07 code/xd.py

@@ -1,0 +1,363 @@
+"""
+XD version originally by Jake Vanderplas/astroML:
+https://github.com/astroML/astroML/blob/master/astroML/density_estimation/xdeconv.py
+
+Modifications by Ross Fadely:
+- 2014-09-23: port and erase astroML dependencies.
+- 2014-09-25: convert class to functions for easy multiprocessing,
+              only Estep seems to be useful across mutliple threads.
+- 2014-09-29: add simple covarince regularization to avoid singular matricies.
+- 2014-11-03: add ability to fix means and align covariances.
+
+Extreme deconvolution solver
+
+This follows Bovy et al.
+http://arxiv.org/pdf/0905.2979v2.pdf
+
+Arbitrary mixing matrices R are not yet implemented: currently, this only
+works with R = I.
+"""
+
+import multiprocessing
+import numpy as np
+
+from time import time
+from sklearn.mixture import GMM
+from utils import logsumexp, log_multivariate_gaussian
+from utils import log_multivariate_gaussian_Nthreads
+from gmm_wrapper import constrained_GMM
+
+def XDGMM(X, Xcov, n_components, n_iter=100, tol=1E-5, Nthreads=1, R=None, 
+          init_n_iter=10, w=None, model=None, fixed_means=None,
+          aligned_covs=None, verbose=False):
+    """
+    Extreme Deconvolution
+
+    Fit an extreme deconvolution (XD) model to the data
+
+    Parameters
+    ----------
+    n_components: integer
+        number of gaussian components to fit to the data
+    n_iter: integer (optional)
+        number of EM iterations to perform (default=100)
+    tol: float (optional)
+        stopping criterion for EM iterations (default=1E-5)
+    X:    array_like
+          Input data. shape = (n_samples, n_features)
+    Xcov: array_like
+          Covariance of input data.  shape = (n_samples, n_features,
+          n_features)
+    R:    array_like
+          (TODO: not implemented)
+          Transformation matrix from underlying to observed data.  If
+          unspecified, then it is assumed to be the identity matrix.
+    w:    float or array_like
+          if float - w * np.eye is added to V
+          if vector - np.diag(w) is added to V
+          if array - w is added to V 
+    Notes
+    -----
+    This implementation follows Bovy et al. arXiv 0905.2979
+    """
[…]

### 2014-11-07 code/utils.py
rm debug statement

@@ -124,14 +124,8 @@ def log_multivariate_gaussian(x, mu, V, Vinv=None, method=1):
 
     elif method == 1:
         if Vinv is None:
-            try:
-                Vinv = np.array([np.linalg.inv(V[i])
-                                 for i in range(V.shape[0])]).reshape(Vshape)
-            except:
-                for i in range(V.shape[0]):
-                    print V[i]
-                    print np.linalg.inv(V[i])
-                    assert 0
+            Vinv = np.array([np.linalg.inv(V[i])
+                             for i in range(V.shape[0])]).reshape(Vshape)
         else:
             assert Vinv.shape == Vshape
 

### 2014-11-07 code/gmm_wrapper.py
more code, some license

@@ -1,0 +1,751 @@
+#
+# Gaussian Mixture Model, swiped from scikit-learn but with 
+# last too functions added to allow fixed means and axis-aligned
+# covariance.  - Ross Fadely
+#
+import numpy as np
+
+"""
+Gaussian Mixture Models.
+
+This implementation corresponds to frequentist (non-Bayesian) formulation
+of Gaussian Mixture Models.
+"""
+
+# Author: Ron Weiss <ronweiss@gmail.com>
+#         Fabian Pedregosa <fabian.pedregosa@inria.fr>
+#         Bertrand Thirion <bertrand.thirion@inria.fr>
+
+import numpy as np
+from scipy import linalg
+
+from sklearn.base import BaseEstimator
+from sklearn.utils import check_random_state
+from sklearn.utils.extmath import logsumexp, pinvh
+from sklearn import cluster
+
+from sklearn.externals.six.moves import zip
+
+EPS = np.finfo(float).eps
+
+
+def log_multivariate_normal_density(X, means, covars, covariance_type='diag'):
+    """Compute the log probability under a multivariate Gaussian distribution.
+
+    Parameters
+    ----------
+    X : array_like, shape (n_samples, n_features)
+        List of n_features-dimensional data points.  Each row corresponds to a
+        single data point.
+    means : array_like, shape (n_components, n_features)
+        List of n_features-dimensional mean vectors for n_components Gaussians.
+        Each row corresponds to a single mean vector.
+    covars : array_like
+        List of n_components covariance parameters for each Gaussian. The shape
+        depends on `covariance_type`:
+            (n_components, n_features)      if 'spherical',
+            (n_features, n_features)    if 'tied',
+            (n_components, n_features)    if 'diag',
+            (n_components, n_features, n_features) if 'full'
+    covariance_type : string
+        Type of the covariance parameters.  Must be one of
+        'spherical', 'tied', 'diag', 'full'.  Defaults to 'diag'.
+
+    Returns
+    -------
+    lpr : array_like, shape (n_samples, n_components)
+        Array containing the log probabilities of each data point in
+        X under each of the n_components multivariate Gaussian distributions.
+    """
+    log_multivariate_normal_density_dict = {
+        'spherical': _log_multivariate_normal_density_spherical,
+        'tied': _log_multivariate_normal_density_tied,
+        'diag': _log_multivariate_normal_density_diag,
+        'full': _log_multivariate_normal_density_full}
+    return log_multivariate_normal_density_dict[covariance_type](
[…]

### 2014-11-07 code/gmm_wrapper.py
typo in comment

@@ -1,6 +1,6 @@
 #
 # Gaussian Mixture Model, swiped from scikit-learn but with 
-# last too functions added to allow fixed means and axis-aligned
+# last two functions added to allow fixed means and axis-aligned
 # covariance.  - Ross Fadely
 #
 import numpy as np

### 2014-11-07 code/gmm_wrapper.py
update comment

@@ -1,7 +1,11 @@
 #
-# Gaussian Mixture Model, swiped from scikit-learn but with 
-# last two functions added to allow fixed means and axis-aligned
-# covariance.  - Ross Fadely
+# Gaussian Mixture Model, swiped from scikit-learn then tweaked -
+# 1) eliminate need do initialization so we can specify a previous set
+#    of parameters.
+# 2) add last two functions to allow fixed means and axis-aligned
+#    covariance.  
+#
+# Tweaks by - Ross Fadely
 #
 import numpy as np
 

### 2014-11-07 code/gmm_wrapper.py
elim double import

@@ -7,8 +7,6 @@
 #
 # Tweaks by - Ross Fadely
 #
-import numpy as np
-
 """
 Gaussian Mixture Models.
 

### 2014-11-07 code/gmm_wrapper.py
comment on the wrap

@@ -713,6 +713,14 @@ def constrained_GMM(X, K, iter_total, fixed_means=None, aligned_covs=None,
     """
     Wrap the sklearn mixture of gaussains.  Take one iter at a time,
     apply the constraint after each step.
+    
+    X: data
+    K: n_components
+    iter_total: number of gmm iterations
+    fixed_means: array of length n_dim, (np.inf, np.inf) if no fix 
+                 otherwise tuple with mean for the component
+    aligned_covs: list of tuples. First element is component number
+                  second is list of feature indices to align.
     """
     assert (fixed_means is not None) | (aligned_covs is not None) | \
         (cov_minmax is not None), '\n\nWhy are you using this code??\n'

### 2014-11-10 code/utils.py
plotting

@@ -214,7 +214,7 @@ def fetch_epoch(epoch, kind, verbose=False):
     Return the single epoch and the matched coadded data.
     """
     assert kind in ['stars', 'gals']
-    ddir = os.environ['sgdata']
+    ddir = os.environ['xddata']
 
     # single epoch
     f = pf.open(ddir + 's82single_%s_%d.fits' % (kind, epoch))
@@ -295,7 +295,7 @@ def fetch_prepped_dr10data(N, fgal=0.5, features=['psf_mag', 'model_colors',
     Prepare SDSS DR10 data to run XD.
     """
     np.random.seed(seed)
-    ddir = os.environ['sgdata']
+    ddir = os.environ['xddata']
     f = pf.open(ddir + 'dr10_30k_stars.fits')
     d = f[1].data
     f.close()
@@ -386,6 +386,33 @@ def prep_data(d, features, filters=None, max_err=100.):
     Xcov = np.zeros(Xerr.shape + Xerr.shape[-1:])
     Xcov[:, range(Xerr.shape[1]), range(Xerr.shape[1])] = Xerr ** 2
     Xcov = np.tensordot(np.dot(Xcov, W.T), W, (-2, -1))
+
+    return X, Xcov
+
+def fetch_glob_data(name, features, filters):
+    """
+    Return the data and cov matrix for SDSS DR10 data for the specified
+    globular cluster.
+    """
+    assert name.lower() in ['m3', 'm5', 'm15']
+    from glob import glob
+    files = glob(os.environ['xddata'] + name.lower() + '*fits')
+
+    f = pf.open(files[0])
+    d = f[1].data
+    d = d[(d['clean'] == 1) & (d['type'] == 6)]
+    X, Xcov = prep_data(d, features, filters)
+    f.close()
+
+    if len(files) > 1:
+        for i in range(1, len(files)):
+            f = pf.open(files[i])
+            d =f[1].data
+            d =d[(d['clean'] == 1) & (d['type'] == 6)]
+            tx, txc = prep_data(d, features, filters)
+            f.close()
+            X = np.vstack((X, tx))
+            Xcov = np.vstack((Xcov, txc))
 
     return X, Xcov
 

### 2014-11-10 code/xd.py

@@ -344,8 +344,8 @@ class xd_model(object):
             iV[i] = np.linalg.inv(self.V[i])
 
         for i in range(Xcov.shape[0]):
-            if i % 100 == 0:
-                print i
+            if (model.verbose & i % 1000):
+                print 'Posterior Calculation for datum %d' % i
             Xicov = np.linalg.inv(Xcov[i])
             for j in range(self.n_components):
                 dlt = X[i] - self.mu[j]

### 2014-11-11 code/utils.py
petro features, alpha limits

@@ -305,8 +305,10 @@ def fetch_prepped_dr10data(N, fgal=0.5, features=['psf_mag', 'model_colors',
     f.close()
     Xgal, Xgalcov = prep_data(d, features, filters)
 
-    Nmax = np.minimum(Xstar.shape[0], Xgal.shape[0])
-    assert N <= Nmax, 'Not enough data for request'
+    Nmax = N * fgal
+    assert Xgal.shape[0] <= Nmax, 'Not enough galaxy data for request'
+    Nmax = N - Nmax
+    assert Xstar.shape[0] <= Nmax, 'Not enough star data for request'
 
     Ngal = np.round(N * fgal).astype(np.int)
     Nstar = N - Ngal
@@ -337,10 +339,12 @@ def make_W_matrix(features, filters, odim):
         # begin spaghetti code
         if 'psf' in feature:
             ind = 0
-        else:
+        elif 'model' in feature:
             ind = 5
+        else:
+            ind = 10
 
-        if 'mag' in feature:
+        if ('mag' in feature) or ('petro' in feature):
             for f in filters[i]:
                 W = np.vstack((W, np.zeros((1, odim))))
                 W[idx, ind + ref[f]] = 1.
@@ -363,7 +367,7 @@ def make_W_matrix(features, filters, odim):
             
     return W[:-1]
 
-def prep_data(d, features, filters=None, max_err=100.):
+def prep_data(d, features, filters=None, max_err=1000., s=0.396):
     """
     Return the prepared data.
     """
@@ -377,9 +381,25 @@ def prep_data(d, features, filters=None, max_err=100.):
                            d['extinction_' + f]for f in 'ugriz']).T
     modelmagerrs = np.vstack([d['modelmagerr_' + f] for f in 'ugriz']).T
 
-    X = np.hstack((psfmags, modelmags))
-    Xerr = np.hstack((psfmagerrs, modelmagerrs))
-    
+    try:
+        seeing = np.vstack([d['mrrcc_' + f] for f in 'ugriz']).T
+        ind = seeing < 0.0
+        seeing = 2. * s * np.sqrt(np.log(2.) * np.abs(seeing))
+        petror50s = np.vstack([d['petror50_' + f] for f in 'ugriz']).T
+        petror50errs = np.vstack([d['petror50err_' + f] for f in 'ugriz']).T
+        ind = ind | (petror50s < 0.) | (np.abs(petror50errs) > max_err)
[…]

### 2014-11-11 code/xd.py

@@ -30,7 +30,7 @@ from gmm_wrapper import constrained_GMM
 
 def XDGMM(X, Xcov, n_components, n_iter=100, tol=1E-5, Nthreads=1, R=None, 
           init_n_iter=10, w=None, model=None, fixed_means=None,
-          aligned_covs=None, verbose=False):
+          aligned_covs=None, seed=None, verbose=False):
     """
     Extreme Deconvolution
 
@@ -61,6 +61,9 @@ def XDGMM(X, Xcov, n_components, n_iter=100, tol=1E-5, Nthreads=1, R=None,
     -----
     This implementation follows Bovy et al. arXiv 0905.2979
     """
+    if seed is not None:
+        np.random.seed(seed)
+
     if model is None:
         model = xd_model(X.shape, n_components, n_iter, tol, w, Nthreads,
                          fixed_means, aligned_covs, verbose)
@@ -198,9 +201,16 @@ def _Mstep(model, q, b, B):
     M-step: compute alpha, mu, V, update to model
     """
     qj = q.sum(0)
+    alpha = qj / model.n_samples
+
+    # prevent no component from having too low a weight
+    ind = alpha < model.min_alpha
+    alpha[ind] = model.min_alpha
+    alpha /= alpha.sum()
+    qj = alpha * model.n_samples
 
     # update alpha
-    model.alpha = qj / model.n_samples
+    model.alpha = alpha
 
     # update mu
     model.mu = np.sum(q[:, :, np.newaxis] * b, 0) / qj[:, np.newaxis]
@@ -232,7 +242,7 @@ class xd_model(object):
     Class to store all things pertinent to the XD model. 
     """
     def __init__(self, xshape, n_components, n_iter, tol, w, Nthreads,
-                 fixed_means, aligned_covs, verbose):
+                 fixed_means, aligned_covs, verbose, min_alpha=1.e-100):
         self.n_samples = xshape[0]
         self.n_features = xshape[1]
         self.n_components = n_components
@@ -242,6 +252,7 @@ class xd_model(object):
         self.fixed_means = fixed_means
         self.aligned_covs = aligned_covs
         self.verbose = verbose
+        self.min_alpha = min_alpha
 
         self.V = None
         self.mu = None

### 2014-11-13 documents/paper/ms_v0.0.tex
doc update

@@
 
-\title{Joint photometric modeling for improved measurements in imaging surveys.}
+\title{Joint photometric modeling for improved measurements in imaging
+surveys.}
 \author{Ross~Fadely\altaffilmark{1} \&
         David~W.~Hogg\altaffilmark{1,2}}
-\altaffiltext{1}{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics, New~York~University, 4~Washington~Place, New~York, NY 10003, USA}
-\altaffiltext{2}{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, 69117 Heidelberg, Germany}
+\altaffiltext{1}{Center~for~Cosmology~and~Particle~Physics,
+Department~of~Physics, New~York~University, 4~Washington~Place, New~York,
+NY 10003, USA}
+\altaffiltext{2}{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, 69117
+Heidelberg, Germany}
 
@@
 
-\begin{itemize}
-\item Mixture of Gaussians where each datum's observational uncertainty is accounted for.
-\item cite bovy and applicable works.
-\item regularization to avoid singularities
-\end{itemize}
-
-\subsection{Posterior estimation}
-
-After inference, the XD mixture provides an estimation for the noiseless distribution of 
-the data:
-
-\begin{eqnarray}\displaystyle
-p(\noiselessdata) = \sum_j^K \frac{\alpha_j}{\sqrt{2\pi\det{|\xdcov|}}} \exp \left (-\frac{1}{2}(\noiselessdata - \vmu_j)^\transpose \xdcov^{-1}(\noiselessdata - \vmu_j) \right )
-\quad .
-\label{eqn:noiseless}
-\end{eqnarray}
-
-In a hierarchical sense, this learned distribution provides a prior distribution on what the 
-value of a datum should be --- the distribution of the denoised data $\noiselessdata$ is 
-learned for all the observations via XD.  Using Bayes' Rule, we can estimate the posterior 
-value for a given datum
+Our goal is to construct a model of the joint distribution of the space of
+possible photometric measurements.  The key idea is that once we have learned
+what the underlying distribution can be, we can utilize this knowledge as a
+prior distribution over what the set of \emph{true} observations should be for
+a given object.  That is, for each datum we want to construct the posterior
+for the true photometric quantities -
 
@@
 
-\noindent where we assume that the prior for $\truedatum$ is $p(\noiselessdata)$.  Since 
-$p(\noiselessdata)$ is a sum of Gaussian densities, and we multiply this by a Gaussian 
-likelihood, the posterior is also a sum of Gaussians.  That is, the posterior for the true 
-value of the observation is
+\noindent where $\datum$ is the observed vector of photometric measurements, 
+$\truedatum$ is the set of values we would have observed in the absense of
+noise, and $\datumcov$ is the covariance for the observation which is
+constructed from our (e.g, the survey's) noise estimates.  The distribution 
+$p(\truedatum)$ is the prior distribution we wish to learn, and the likelihood
+$p(\datum | \truedatum)$ is assumed to be Gaussian.  Since we are learning the
+prior distribution from a set of many observations, this is refered to as a
+multilevel or hierarchical model.
+
+The art in building a hierachical model is deciding the structure of the
+distributions involved -- how distributions are parameterized, what is held
+fixed or considered know, and what the relationships are between the various
+distributions.  We believe the distribution of $p(\truedatum)$ can be quite 
+complex, and it is not obvious what the right description of the distribution
+might be.  For these reasons, we choose to use Extreme Deconvolution
+
+After inference, the XD mixture provides an estimation for the noiseless
+distribution of the data:
+
+\begin{eqnarray}\displaystyle
+p(\noiselessdata) = \sum_j^K \frac{\alpha_j}{\sqrt{2\pi\det{|\xdcov|}}} \exp \left (-\frac{1}{2}(\noiselessdata - \vmu_j)^\transpose \xdcov^{-1}(\noiselessdata - \vmu_j) \right )
+\quad .
+\label{eqn:noiseless}
+\end{eqnarray}
+
+In a hierarchical sense, this learned distribution provides a prior
+distribution on what the value of a datum should be --- the distribution of
+the denoised data $\noiselessdata$ is learned for all the observations via XD.
+Using Bayes' Rule, we can estimate the posterior value for a given datum
+
+\noindent where we assume that the prior for $\truedatum$ is
+$p(\noiselessdata)$.  Since $p(\noiselessdata)$ is a sum of Gaussian
+densities, and we multiply this by a Gaussian likelihood, the posterior is
+also a sum of Gaussians.  That is, the posterior for the true value of the
+observation is
 

### 2014-11-13 documents/paper/ms_v0.0.tex
more text

@@
 \newcommand{\transpose}{{\rm T}}
+\newcommand\ttt[1]{{\texttt{#1}}}
 \newcommand\rf[1]{{\bf [RF: #1]}}
@@
 complex, and it is not obvious what the right description of the distribution
-might be.  For these reasons, we choose to use Extreme Deconvolution
+might be.  For these reasons, we choose to use Extreme Deconvolution (XD)
+\citep{bovy09}.  XD is a simple but powerful and flexible model which models
+the density of the data as a Mixture of Gaussians, accounting for the 
+measurement errors of the observations.  In short, XD learns a model for which 
+each datum is drawn from an underlying distribution which is convolved by the 
+specified noise estimates.
 
-After inference, the XD mixture provides an estimation for the noiseless
-distribution of the data:
+After inference, the XD mixture provides an estimation for the `noiseless' (or
+less noisy) distribution of the data:
 
@@
 
-In a hierarchical sense, this learned distribution provides a prior
-distribution on what the value of a datum should be --- the distribution of
-the denoised data $\noiselessdata$ is learned for all the observations via XD.
-Using Bayes' Rule, we can estimate the posterior value for a given datum
+\noindent From here, we assume that the prior for $\truedatum$ is
+$p(\noiselessdata)$.  Doing so, we consider the analysis here akin to an 
+`empirical' heirarchical model -- the prior is not learned simultaneously with
+the posterior but rather is learned separately from the collection of
+observations.
 
-\noindent where we assume that the prior for $\truedatum$ is
-$p(\noiselessdata)$.  Since $p(\noiselessdata)$ is a sum of Gaussian
-densities, and we multiply this by a Gaussian likelihood, the posterior is
-also a sum of Gaussians.  That is, the posterior for the true value of the
-observation is
+Since $p(\noiselessdata)$ is a sum of Gaussian densities, and we multiply this
+by a Gaussian likelihood, the posterior is also a sum of Gaussians.  That is,
+the posterior for the true value of the observation is
 
@@
 \end{eqnarray}
+
+\section{Data and Model Considerations}
+
+To demonstrate the utility of our approach, we focus on the $ugriz$
+photometric data provided by SDSS' Data Release 10 \citep[DR10,][]{?} and
+Stripe 82 \citep[S82,][]{annis14, others?}.  Here, our primary goal is to
+provide better estimate of the dereddened PSF and model magnitudes observed by
+SDSS.  More specifically, we aim to model the distribution of the $r$ band PSF
+magnitudes, the $u-g$, $g-r$, $r-i$, and $i-z$ model colors, and the different
+between PSF and model magnitudes in the $ugriz$ bands.  Thus, are fiducial 
+models aim to understand the distribution of 10 measurements or `features' per 
+observed source.
+
+In DR10, we randomly select 30k `stars' and `galaxies' (\ttt{type} $= 6, 3$)
+which PSF $r$ magnitudes between 18 and 22, which are \ttt{Primary}
+photometric observations, and which have clean
+photometry\footnote{Observations which are not true in \ttt{EDGE},
+\ttt{NOPROFILE}, \ttt{PEAKCENTER}, \ttt{NOTCHECKED}, \ttt{PSF\_FLUX\_INTERP},
+\ttt{SATURATED}, \ttt{BAD\_COUNTS\_ERROR}, \ttt{DEBLEND\_NOPEAK},
+\ttt{INTERP\_CENTER}, or \ttt{COSMIC\_RAY}, and is detected in
+\ttt{BINNNED1}.}.  The even number of stars and galaxies selected for our
+sample is to mitigate any biases we might induce by favoring one population
+over the other.  For purpose of comparing lower signal-to-noise obversations
+to ones of higher quality, we also select a similar set of observations from
+the S82 coadd \citep{annis14}.  S82 is unique in the SDSS dataset as it is one
+of the few region of the sky for which on average $\sim 20$ epochs of imaging
+have been taken.  As a result, the median $r$ PSF magnitude uncertainty in a
+single epoch is BLAH compared to a median of BLAH for the coadded data (c.f, 
+\ref{fig:error_rates}).  For the results presented below, we have trained our
+XD model on the DR10 data described above and use this model to calculate
+posteriors for both DR10 and S82 single epoch data.
+
+For our XD model, we must specify the number of components in the XD mixture.
+Since our goal is simply to demonstrate the utility of our method, we set the 
+number somewhat arbitrarily to 32.  We also tweak the XD algorithm presented in
+\citet{bovy09} by allowing the user to specify the mean value of a feature,
+as well as inforce that the covariance for a feature to be axis aligned
+(diagonal elements only).  For PSF minus model magnitude features, we force
+the mean to be zero and for the covariance to be axis aligned since this is
+expected for unresolved sources. Since there are also extended sources in the
+data, we only do so for $N_{\rm point}$ components of the mixture.  Again,
+somewhat arbitrarily we set this number to 16.  We have set the number of
+components in the mixture (both in total and for unresolved sources) to
+sensible numbers given the amount of data used, and seem to have adequate
+performance for our purposes.  If this was done in a science quality or
+survey data product setting, we recommend setting these parameters via cross
+validation with data at higher signal to noise.  Finally we note that we 
+apply a small amount of regularization to the covariance of the mixture
+components, exactly the same in nature to that of \citet{bovy10}.  In short, 
+a small value $w$ is added to the diagonal of the covariances to prevent
+singular matrices during the learning process.  We set $w$ to be 1\% of the 
+smallest uncertainty measured in each feature, and find that this does not
+significantly affect our final results.
+
+\section{Results and Discussion}
+
+We find the XD model provides a good understanding of the distribution of the 
+data, and that the posteriors we compute are sensible, less noisy estimates of 
+the observations.  To start, we examine the distribution of the XD mixture 
+relative to the DR10 data.  In figure \ref{fig:contours} we show the contours 
+of the $N_{\rm point}$ components along with data labeled as `star' in DR10,
+and likewise for extended sources.  While its difficult to assess the quality
+of the XD model from the plots, there are a few supporting points.  For the 
+unresolved sources, the mixture prefers low variance in the value of $r$ PSF
+minus model magnitudes which is in agreement we our intuition that stars
+ought to be clustered around values of zero.  Second, the stellar locus in
+color is clear and well tracked by the XD $N_{\rm point}$ components.  Though 
+galaxies have less defining characteristics, it is clear that the XD mixture 
+also tracks their distribution well.
+
+In figure \ref{fig:posteriors}, we show the posterior distributions for 3
+stars and 3 galaxies at different magnitudes.
+
+So far we have shown operationally that the XD model does sensible modeling of
+the photometry and is more \emph{precise}, but it is important to ask whether
+the posteriors are at all \emph{accurate}.  To do so, we rely on comparsions of
+the posteriors we get for S82 single epoch data versus the data from the
+coadd.  In figure \ref{fig:xx_plots}, we plot the different between the single
+epoch and coadd data, as well as our XD posteriors versus the coadd.  We find
+that the median of the residuals for our XD posteriors is consistent with
+zero, indicating the model is not only precise but also accurate.  We note
+this is true for all of the features we model.  Also obvious is the clear
+improvement in the photometry over the single epoch data.
+
+In SDSS, star-galaxy classification is accomplished by selecting objects a
+small range of PSF minus model magnitudes and calling them stars.  In DR10,
+the range is approximately\footnote{The actual algorith involves a combination
+of this criterion for $gri$ bands.} $|r_{PSF} - r_{model}| < 0.145$ while the
+S82 coadd criterion is $|r_{PSF} - r_{model}| < 0.03$.  In figure \ref{fig:pmm}
+we show a normalized histogram of $r$ PSF minus model magnitudes around
+$r=21$, for objects that are called stars in the coadd.  Note that our XD
+posteriors dramatically tighten the distribution of objects around near zero,
+and also indicate the present of interloping galaxies.
+
 \end{document}

### 2014-11-14 documents/paper/ms_v0.0.tex
fig text

@@
 
+\begin{figure}
+\centering
+\includegraphics[clip=true, trim=1.5cm 0.5cm 1.5cm 0.5cm,
+  width=16cm]{fig1.png}
+\caption{Shown are three different projections of one tenth of the DR10
+photometric data used (black points), with the top/bottom row corresponding
+objects classified as stars/galaxies (\ttt{type} $=6, 3$).  Gray contours
+indicate the different XD components.  For the top we only plot the
+$N_{\rm point}$ components while in the bottom we plot the rest.  Note,
+however, when we compute our model we do not use SDSS classifications -- all
+data is modeled with the combination of both top and bottom contours.  Note
+the components modeling psfmag minus modelmag for starlike objects prefer
+small variance, and generally the model follows the data well.
+}
+\label{fig:contours}
+\end{figure}
+
 In figure \ref{fig:posteriors}, we show the posterior distributions for 3

### 2014-11-18 documents/paper/ms_v0.0.tex
more figs/text

@@
 
-In figure \ref{fig:posteriors}, we show the posterior distributions for 3
-stars and 3 galaxies at different magnitudes.
+\begin{figure}
+\centering
+\includegraphics[clip=true, trim=1.5cm 0.5cm 1.5cm 0.5cm,
+  width=16cm]{fig2.png}
+\caption{Error ellipses for single epoch S82 photometry (dashed) and XD
+posterior distributions for a star (top row) and galaxy (bottom row).  The 
+projections are the same as in Figure \ref{fig:contours}.  Ellipses for the XD
+posteriors are generally smaller, yet are still consistent with the original 
+measurements.
+}
+\label{fig:posteriors}
+\end{figure}
+
+In figure \ref{fig:posteriors}, we show the posterior distributions for a
+random star and galaxy in our sample.  We find generally that projections of 
+the posterior are consistent with the original measurements, but are of higher 
+precision.  Figure \ref{fig:error_rates} quantifies this, showing the median 
+uncertainty as a function of $r$ magnitude.  We find that the median
+uncertainties of our XD posteriors are a significant improvement to the
+catalog data.  For instance, at $r\sim20$, the uncertainty in the DR10 $r$ PSF 
+magnitude is reduced from 0.28 to 0.2 \rf{DOUBLE CHECK} and more dramatically
+the error in the $u-g$ model color is reduced from 0.5 to 0.3.  It is not
+surprising that the improvements are most significant where the data is more 
+noisy since this is where an informative prior can make the most difference.
+
+\begin{figure}
+\centering
+\includegraphics[clip=true, trim=1.5cm 0.5cm 1.5cm 0.5cm,
+  width=16cm]{fig3.png}
+\caption{Plotted are median values for the 1 sigma uncertainty in $r$ PSF
+magnitude, $r$
+psf minus model magnitude, and the four model colors used.  The solid black 
+curve is for the Stripe 82 coadd, while the dotted black and red curves are for
+single epoch Stripe 82 and DR10 data, respectively.  The dashed curves show the
+median uncertainty for the XD posteriors.  While not quite as precise as the 
+coadd data, the XD posteriors are a significant improvement over the catalog
+data.
+}
+\label{fig:error_rates}
+\end{figure}
 
@@
 the posteriors we get for S82 single epoch data versus the data from the
-coadd.  In figure \ref{fig:xx_plots}, we plot the different between the single
-epoch and coadd data, as well as our XD posteriors versus the coadd.  We find
-that the median of the residuals for our XD posteriors is consistent with
-zero, indicating the model is not only precise but also accurate.  We note
-this is true for all of the features we model.  Also obvious is the clear
-improvement in the photometry over the single epoch data.
+coadd.  In figure \ref{fig:xx_plots}, we plot the single
+epoch data against coadd measurements, as well as our XD posteriors versus the
+coadd.  We find that the XD posteriors are in good agreement with the coadd
+data -- there is no obvious biases or deviations appart from scatter due to
+noise.  We note that there is a slight offset to lower values of $r$ PSF
+magnitudes at the faintest end ($r \gtrsim 21.5$).  This is an artifact due to
+the fact that none of the components we used were constrained to have mean
+around or greater than $r \sim 21.5$.  As a result, the posteriors will
+naturally pull the magnitude values down.  A simple fix is to force a few
+components of the XD mixture to have means at faint magnitude values, but given
+the size of this effect we have not done so for the purposes here.  Note a
+similar effect will also occur at the bright end of our sample, but is less 
+significant since the measurement uncertainties are lower.  As a additional 
+demonstration of the reasonability of our method, we show the color-magnitude
+diagram for DR10 data of the globular cluster Messier 3 in Figure
+\ref{fig:m3}.  Compared to the DR10 data the XD posterior is clearly much
+cleaner at faint $r$ magnitudes, and is more concentrated across much of the
+main sequence.  Note that the width of the main sequence around $r \sim 19$ is
+not dramatically changed from DR10.  This is because the measurement
+uncertainties are fairly small so the DR10 value is the dominant contributor
+to the posterior.  In other words, the XD values seem to be respectful of the 
+data and do not seem to overly `denoise'.
+
+\begin{figure}
+\centering
+\includegraphics[clip=true, trim=1.5cm 0.5cm 1.5cm 0.5cm,
+  width=16cm]{fig4.png}
+\caption{In the top row, we plot 2D histograms of $r$ PSF magnitudes, $u-g$
+model color, and $i-z$ model color for single epoch Stripe 82 data versus the
+corresponding coadd values.  In the bottow row, we plot the same for a single
+sample from the XD posteriors of the data in the top.  Generally speaking the
+XD posteriors are very consistent with the coadd photometry, indicating that 
+the models are indeed acurate.
+}
+\label{fig:xx_plots}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=16cm]{fig5.png}
+\caption{The $g-r$ versus $r$ color-magnitude diagram from the outskirts of
+the globular cluster Messier 3.  The left panel shows the original DR10
+measurements and the right shows the median value from the XD posteriors.  Note
+the main sequence is generally more concentrated, especially at faint
+magnitudes.
+}
+\label{fig:m3}
+\end{figure}
 
@@
 small range of PSF minus model magnitudes and calling them stars.  In DR10,
-the range is approximately\footnote{The actual algorith involves a combination
+the range is approximately\footnote{The actual algorithm involves a combination
 of this criterion for $gri$ bands.} $|r_{PSF} - r_{model}| < 0.145$ while the
 S82 coadd criterion is $|r_{PSF} - r_{model}| < 0.03$.  In figure \ref{fig:pmm}
-we show a normalized histogram of $r$ PSF minus model magnitudes around
-$r=21$, for objects that are called stars in the coadd.  Note that our XD
-posteriors dramatically tighten the distribution of objects around near zero,
-and also indicate the present of interloping galaxies.
+we show a histogram of $r$ PSF minus model magnitudes around $r=19.5$
+20.5, and 21.5 for objects that are called `stars' in the coadd.  Note that
+our XD posteriors dramatically tighten the distribution of objects to values
+very close to zero.  This implies that the quality of the star-galaxy
+classification could be substantially improved by selecting on values produced
+by the XD posteriors.
+
+\begin{figure}
+\centering
+\includegraphics[width=16cm]{fig6.png}
+\caption{Histograms of the values of the $r$ PSF minus model magnitudes for 
+objects within 0.15 mag of $r = 19.5$, 20.5, and 21.5 that were labeled as 
+`stars' in the Stripe 82 coadd.  The dotted histograms shows the distribution
+for single epoch Stripe 82 data, while the solid histograms show those for the
+XD posteriors.  The XD values are strongly peaked near zero, while the single
+epoch values are more distributed, suggesting star-galaxy classification
+could be improved by relying on the XD values.  
+}
+\label{fig:pmm}
+\end{figure}
 

### 2014-11-18 documents/paper/ms_v0.0.tex
more text

@@
 
+We emphasize that the method we present here is just a toy example of how one 
+might improve photometric measurements in astronomical surveys.  By no means do
+we believe that this is the best or only way to build a hierarchical model for 
+the photometry, however, it does have the property of being simple,
+interpretable, and yet quite flexible.  For a ``production'' level set of
+posteriors there would still be much to sort out, including ---
+
+\begin{itemize}
+\item What data should be used for training the XD model?
+\item What are the optimal number of XD components?
+\item Which mean values of the mixture should be fixed
+\item Which covariances should be diagonal?
+\item What are the best features to include model?
+\item What, if any, features should be included that are outside the survey
+data?
+\end{itemize}
+
+All such questions would be straightforward to answer using cross validation,
+either on left out ``test'' data or higher signal-to-noise data.  One of the
+main reasons that we have not done so here is that the correct metric for any
+cross validation approach depends on the scientific goals of the experiment.  
+Large surveys can likely define a set of such metrics that meet much of their 
+users goals, but there may still be cases where individuals may want to define
+their own models.  Finally, we do not argue that posterior values should 
+replace photometric measurements in all cases.  Indeed, in the case transient
+or varying objects this would likely be a bad idea.  Additionally, it may
+hinder the discovery of rare types of objects since the prior would drive the
+measurments to values like more common sources.
+
 \end{document}

### 2014-12-02 documents/paper/ms_v0.0.tex
some intro; pseudocode

@@
+\newcommand{\xdmean}{\datavector{\mu}}
 \documentclass[12pt,preprint]{aastex}
 \usepackage{amssymb,amsmath,mathrsfs,hyperref,datetime}
+
+\usepackage{algorithmic,algorithm}
+\usepackage{eqparbox}
+\renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENT}
+                                      {\it~#1}}
+\newcommand{\blankcomment}{\protect\phantom{$1$}\hspace{10cm}
+                           \protect\phantom{$1$}}
 
@@
 \newcommand{\noiselessdatum}{\tilde{\datum}}
+\newcommand{\xij}{\datavector{x}_{ij}}
+\newcommand{\Lij}{\datavector{\lambda}_{ij}}
+\newcommand{\Tij}{\datavector{T}_{ij}}
+\newcommand{\qij}{\datavector{q}_{ij}}
+\newcommand{\qj}{\datavector{q}_{j}}
+\newcommand{\dlt}{(\xdmu_j - \xij)}
 \newcommand{\xdcov}{\datavector{V}_j}
@@
 \newcommand{\postcov}{\tilde{\datacov}_{i, j}}
-\newcommand{\vmu}{\datavector{\mu}}
-\newcommand{\postmu}{\tilde{\vmu}_{i, j}}
+\newcommand{\xdmu}{\datavector{\mu}}
+\newcommand{\fixedmu}{\bar{\mu}}
+\newcommand{\postmu}{\tilde{\xdmu}_{i, j}}
 \newcommand{\datumcov}{\datacov_i}
-\newcommand{\transpose}{{\rm T}}
+\newcommand{\transpose}{\mathsf{T}}
 \newcommand\ttt[1]{{\texttt{#1}}}
@@
 
+For a large catagory of survey missions, the main scientific deliverable is a
+catalog of photometric measurements such as astrometry, magnitudes, and
+morphological information.  These catalogs are almost always constructed on a 
+per-object basis -- each source in an image is measured 
+independently\footnote{Apart from issues associated with blending, detection,
+psf modeling, etc.}.
+For most types of sources (stars, galaxys and quasars), we have thousands to
+millions of measurements albeit with varying properties and signal-to-noise.
+This begs the question as to whether we can make a better estimate of a given
+photometric measurement based on the fact that we have many similar
+measurements.  Indeed, if we have thousands of measurements of (for example)
+main sequence F4 stars, we ought to be able to get a better estimate of a new 
+ones photometric properties by incorperating what is learned from the
+population.
+
+Since telescope time is expensive both in terms of cost and person-hours, it
+behoves us to consider whether we can make better estimates of photometric 
+measurements at (essentially) fixed expense.  We argue that it is indeed
+possible to improve the measurements, by exploiting the similarity across
+observations.  As a demonstration, we use Extreme Deconvolution
+\citep[XD, ][]{bovy09} to model the photometry in the SDSS DR10 \citep{} and
+Stripe 82 \citep{} data releases.  XD has been shown to be a powerful yet
+simple way to model the properties of heteroskedastic data \citep{bovy10}.
+Here, we apply it to SDSS photometry and produce posterior estimates of the 
+measurements.  In Section \ref{sec:method} we discuss the ideas and methods 
+behind our modeling, Section \ref{sec:data} describes the data we use for 
+our demostration, and \ref{sec:results} presents and discusses the application
+of our XD model.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Extreme Deconvolution}
+\section{Posterior Estimates with Extreme Deconvolution}
+\label{sec:method}
 
@@
 
+\begin{algorithm}
+\caption{Expectation-Maximization for XD}
+\textbf{Expectation Step:}
+\begin{algorithmic}[1]
+\STATE $\Tij \gets \xdcov + \datumcov$
+\STATE $\qij \gets \frac{\alpha_j \mathcal{N}(\datum | \xdmu_j, \Tij)}
+       {\sum_j \alpha_j \mathcal{N}(\datum | \xdmu_j, \Tij)}$
+\STATE $\xij \gets \xdmu_j + \xdcov\Tij^{-1}(\datum - \xdmu_j)$
+\STATE $\Lij \gets \xdcov - \xdcov\Tij^{-1}\xdcov$
+\end{algorithmic}
+\textbf{Maximization Step:}
+\begin{algorithmic}[1]
+\STATE $\alpha_j \gets \frac{1}{N}\sum_i \qij$
+\STATE $\xdmu_j \gets \frac{1}{\qj}\sum_i \qij\xij$
+\FOR [\blankcomment]{$k = 1, \ldots, N_{\rm dim}$}
+    \IF [// fix mean of feature k to $\fixedmu_{jk}$ if 
+        desired]{$\exists\, \fixedmu_{jk}$}
+        \STATE $(\xdmu_j)_k \gets \fixedmu_{jk}$
+    \ENDIF
+\ENDFOR
+\STATE $\xdcov \gets \frac{1}{\qj}\sum_i \qij [\dlt\dlt^\transpose + \Lij]$
+\FOR {$k = 1, \ldots, N_{\rm dim}$}
+    \IF [// axis align covariance for feature k if desired]
+        {$align_{jk}$}
+        \STATE $var = (\xdcov)_{k,k}$
+        \STATE $(\xdcov)_{k,:} \gets 0$
+        \STATE $(\xdcov)_{:,k} \gets 0$
+        \STATE $(\xdcov)_{k,k} \gets var$
+    \ENDIF
+\ENDFOR
+\end{algorithmic}
+\end{algorithm}
+
 After inference, the XD mixture provides an estimation for the `noiseless' (or
@@
 \begin{eqnarray}\displaystyle
-p(\noiselessdata) = \sum_j^K \frac{\alpha_j}{\sqrt{2\pi\det{|\xdcov|}}} \exp \left (-\frac{1}{2}(\noiselessdata - \vmu_j)^\transpose \xdcov^{-1}(\noiselessdata - \vmu_j) \right )
+p(\noiselessdata) = \sum_j^K \frac{\alpha_j}{\sqrt{2\pi\det{|\xdcov|}}} \exp \left (-\frac{1}{2}(\noiselessdata - \xdmu_j)^\transpose \xdcov^{-1}(\noiselessdata - \xdmu_j) \right )
 \quad .
@@
 \postcov & = & (\datumcov^{-1} + \xdcov^{-1})^{-1} \quad {\rm and,} \\
-\postmu & = &  \postcov^{-1} (\datumcov^{-1} \datum + \xdcov^{-1} \vmu_j)
+\postmu & = &  \postcov^{-1} (\datumcov^{-1} \datum + \xdcov^{-1} \xdmu_j)
 \label{eqn:posterior}
@@
 \section{Data and Model Considerations}
+\label{sec:data}
 
@@
 \section{Results and Discussion}
+\label{sec:results}
 
@@
 \centering
-\includegraphics[clip=true, trim=1.5cm 0.5cm 1.5cm 0.5cm,
-  width=16cm]{fig2.png}
+\includegraphics[width=16cm]{fig2.png}
 \caption{Error ellipses for single epoch S82 photometry (dashed) and XD
@@
 \centering
-\includegraphics[clip=true, trim=1.5cm 0.5cm 1.5cm 0.5cm,
-  width=16cm]{fig3.png}
+\includegraphics[width=16cm]{fig3.png}
 \caption{Plotted are median values for the 1 sigma uncertainty in $r$ PSF
@@
 \centering
-\includegraphics[clip=true, trim=1.5cm 0.5cm 1.5cm 0.5cm,
-  width=16cm]{fig4.png}
+\includegraphics[width=16cm]{fig4.png}
 \caption{In the top row, we plot 2D histograms of $r$ PSF magnitudes, $u-g$

### 2014-12-03 documents/paper/ms_v0.0.tex
intro and methods text

@@
 Stripe 82.  Using single epoch data, we show that our joint XD model makes
-predictions that more closely resemble the coadded Stripe 82 data.  Looking
-at the difference between the psf and model magnitudes in the single epoch and 
-coadded data, the average error is XX at $r\sim21$ versus an average of XX
-for our XD posteriors.  In the case of the CMDs of globular clusters, we show
-that the XD posteriors lessen the presence of misclassified galaxies and
-present a narrower stellar locus.  We suggest a number of improvements that
-can be made to provide improved photometric inferences.
+predictions that more closely resemble the coadded Stripe 82 data.  The
+posterior variance of the photometric quantities are generally improved by
+$\sim 50 - 100\%$, with little to no bias relative to the coadded data.
+Looking at the difference between the psf and model magnitudes in the single
+epoch data, the average error is 0.11 mag at $r=21$ versus an average of 0.055
+for our XD posteriors implying a potential dramatic improvement in traditional
+star-galaxy classification. 
 \end{abstract}
@@
 main sequence F4 stars, we ought to be able to get a better estimate of a new 
-ones photometric properties by incorperating what is learned from the
+one's photometric properties by incorperating what is learned from the
 population.
@@
 observations.  As a demonstration, we use Extreme Deconvolution
-\citep[XD, ][]{bovy09} to model the photometry in the SDSS DR10 \citep{} and
-Stripe 82 \citep{} data releases.  XD has been shown to be a powerful yet
+\citep[XD, ][]{bovy11} to model the photometry in the SDSS DR10 \citep{ahn14}
+and
+Stripe 82 \citep{annis14} data releases.  XD has been shown to be a powerful yet
 simple way to model the properties of heteroskedastic data \citep{bovy10}.
@@
 noise, and $\datumcov$ is the covariance for the observation which is
-constructed from our (e.g, the survey's) noise estimates.  The distribution 
+constructed from our (the survey's) noise estimates.  The distribution 
 $p(\truedatum)$ is the prior distribution we wish to learn, and the likelihood
@@
 distributions involved -- how distributions are parameterized, what is held
-fixed or considered know, and what the relationships are between the various
+fixed or considered known, and what the relationships are between the various
 distributions.  We believe the distribution of $p(\truedatum)$ can be quite 
@@
 might be.  For these reasons, we choose to use Extreme Deconvolution (XD)
-\citep{bovy09}.  XD is a simple but powerful and flexible model which models
+\citep{bovy11}.  XD is a simple but powerful and flexible model which models
 the density of the data as a Mixture of Gaussians, accounting for the 
@@
 \caption{Expectation-Maximization for XD}
+\label{alg:em}
 \textbf{Expectation Step:}
@@
 \end{algorithmic}
+\textbf{Note - } $(\mathbf{A})_l$ denotes the $l$-th element of vector
+$\mathbf{A}$, and $(\mathbf{B})_{l,m}$ denotes the element of the $l$-th row
+and $m$-th column of matrix $\mathbf{B}$.  The symbol $:$ indicates all of the
+elements within a row/column/entry.
 \end{algorithm}
 
-After inference, the XD mixture provides an estimation for the `noiseless' (or
-less noisy) distribution of the data:
+The XD mixture provides an estimation for the `noiseless' (or less noisy)
+distribution of the data:
 
 \begin{eqnarray}\displaystyle
-p(\noiselessdata) = \sum_j^K \frac{\alpha_j}{\sqrt{2\pi\det{|\xdcov|}}} \exp \left (-\frac{1}{2}(\noiselessdata - \xdmu_j)^\transpose \xdcov^{-1}(\noiselessdata - \xdmu_j) \right )
+p(\noiselessdata) = \sum_j^J \frac{\alpha_j}{\sqrt{2\pi\det{|\xdcov|}}} \exp \left (-\frac{1}{2}(\noiselessdata - \xdmu_j)^\transpose \xdcov^{-1}(\noiselessdata - \xdmu_j) \right )
 \quad .
@@
 
-\noindent From here, we assume that the prior for $\truedatum$ is
-$p(\noiselessdata)$.  Doing so, we consider the analysis here akin to an 
+\noindent The probability of the intrinsic data values $p(\noiselessdata)$ is
+a sum over $J$ gaussians each with a given mean $\xdmu_j$ and covariance
+$\xdcov$, with weights $\alpha_j$.  An efficient way
+to optimize this model is to use Expectation Maximization (EM) as presented in
+\citep{bovy11}.  We summarize the steps of the EM procedure in Algorithm
+\ref{alg:em}.  XD models probability of each datum as a
+sum of $J$ normals with $\mathcal{N}(\datum | \xdmu_j, \Tij)$.  In the
+expectation step we build $\Tij$ from the component's variance and the datum's
+noise estimation.  Then the conditional distribution of the true value
+$\noiselessdata_i$ is estimated as having mean $\xij$ and covariance
+$\Lij$, with probability $\qij$ (relative across components).  In the
+maximization step, we update the component weights as a average over all
+the $\qij$'s for the data.  The component mean is set by the weighted sum of 
+the $\xij$ values, and the component variance is the weighted sum of the 
+covariance of the mean relative to the $\xij$ plus $\Lij$.  In addition, we
+show in maximization steps $3-7$ and $9-16$ an effective way to fix the mean
+and axis-align part or all of the component, respectively.  We discuss why we
+wish to do so in Section \ref{sec:data}.
+
+We assume that the prior for $\truedatum$ is $p(\noiselessdata)$
+(Eq. \ref{eqn:noiseless}).  Doing so, we consider the analysis here akin to an 
 `empirical' heirarchical model -- the prior is not learned simultaneously with
 the posterior but rather is learned separately from the collection of
-observations.
-
-Since $p(\noiselessdata)$ is a sum of Gaussian densities, and we multiply this
-by a Gaussian likelihood, the posterior is also a sum of Gaussians.  That is,
-the posterior for the true value of the observation is
+observations. Since $p(\noiselessdata)$ is a sum of Gaussian densities, and
+we multiply this by a Gaussian likelihood, the posterior is also a sum of
+Gaussians.  That is, the posterior for the true value of the observation is
 
 \begin{eqnarray}\displaystyle
-p(\truedatum | \datum) & \propto & \sum_j^K \frac{\alpha_j}{\sqrt{2\pi\det{|\postcov|}}} \exp \left (-\frac{1}{2}(\truedatum - \postmu)^\transpose \postcov^{-1}(\truedatum - \postmu) \right ) \\
+p(\truedatum | \datum) & \propto & \sum_j^J \frac{\alpha_j}{\sqrt{2\pi\det{|\postcov|}}} \exp \left (-\frac{1}{2}(\truedatum - \postmu)^\transpose \postcov^{-1}(\truedatum - \postmu) \right ) \\
 {\rm where,}\\
@@
 To demonstrate the utility of our approach, we focus on the $ugriz$
-photometric data provided by SDSS' Data Release 10 \citep[DR10,][]{?} and
-Stripe 82 \citep[S82,][]{annis14, others?}.  Here, our primary goal is to
+photometric data provided by SDSS' Data Release 10 \citep[DR10,][]{ahn14} and
+Stripe 82 \citep[S82,][]{annis14}.  Here, our primary goal is to
 provide better estimate of the dereddened PSF and model magnitudes observed by
@@
 have been taken.  As a result, the median $r$ PSF magnitude uncertainty in a
-single epoch is BLAH compared to a median of BLAH for the coadded data (c.f, 
+single epoch is 0.07 at $r=21$ compared to a median of 0.015 for the coadded
+data (compare to Figure 
 \ref{fig:error_rates}).  For the results presented below, we have trained our
@@
 Since our goal is simply to demonstrate the utility of our method, we set the 
-number somewhat arbitrarily to 32.  We also tweak the XD algorithm presented in
-\citet{bovy09} by allowing the user to specify the mean value of a feature,
+number somewhat arbitrarily to 32.  
+As stated in Section \ref{sec:method} we also tweak the XD algorithm
+presented in
+\citet{bovy11} by allowing the user to specify the mean value of a feature,
 as well as inforce that the covariance for a feature to be axis aligned
@@
 apply a small amount of regularization to the covariance of the mixture
-components, exactly the same in nature to that of \citet{bovy10}.  In short, 
+components, exactly the same in nature to that of \citet{bovy09}.  In short, 
 a small value $w$ is added to the diagonal of the covariances to prevent
@@
 objects classified as stars/galaxies (\ttt{type} $=6, 3$).  Gray contours
-indicate the different XD components.  For the top we only plot the
-$N_{\rm point}$ components while in the bottom we plot the rest.  Note,
-however, when we compute our model we do not use SDSS classifications -- all
-data is modeled with the combination of both top and bottom contours.  Note
+indicate the different XD components.  In the top panel we show the data and 
+our XD mixture.  For the middle we only plot the
+$N_{\rm point}$ components while in the bottom we plot the rest and restrict
+the data to only SDSS point sources.  In the bottom, we plot the remaining
+components and SDSS extended sources.
+Note,
+however, when we compute our model we do not use SDSS classifications -- the 
+top panel is what is modeled.  Interestingly,
 the components modeling psfmag minus modelmag for starlike objects prefer
-small variance, and generally the model follows the data well.
+small variance, in agreement with what is expected.
 }
@@
 \caption{Error ellipses for single epoch S82 photometry (dashed) and XD
-posterior distributions for a star (top row) and galaxy (bottom row).  The 
+posterior distributions for a star (top row) and galaxy (bottom row).  Mean
+values correspond to the cross and square for the S82 and XD photometry, 
+respectively.  The 
 projections are the same as in Figure \ref{fig:contours}.  Ellipses for the XD
@@
 S82 coadd criterion is $|r_{PSF} - r_{model}| < 0.03$.  In figure \ref{fig:pmm}
-we show a histogram of $r$ PSF minus model magnitudes around $r=19.5$
+we show a histogram of $r$ PSF minus model magnitudes around $r=19.5$,
 20.5, and 21.5 for objects that are called `stars' in the coadd.  Note that
@@
 the photometry, however, it does have the property of being simple,
-interpretable, and yet quite flexible.  For a ``production'' level set of
+interpretable, and yet quite flexible.  For a ``production level'' set of
 posteriors there would still be much to sort out, including ---
@@
 \item What are the optimal number of XD components?
-\item Which mean values of the mixture should be fixed
+\item Which mean values of the mixture should be fixed?
 \item Which covariances should be diagonal?
-\item What are the best features to include model?
+\item What are the best features to include in the model?
 \item What, if any, features should be included that are outside the survey
@@
 
+\clearpage
+
+\bibliographystyle{apj}
+\begin{thebibliography}{1}
+
+\bibitem[{{Ahn} {et~al.}}(2014)]{ahn14}
+{{Ahn}, C.~P.,  {Alexandroff}, R.,  {Allende Prieto}, C.,  
+  {Anders}, F.,  {Anderson}, S.~F.,  {Anderton}, T.,  {Andrews},
+  B.~H.,  {Aubourg}, {\'E}.,  {Bailey}, S.,  {Bastien}, F.~A. \&  et al.}
+  2014, \apj, 211, 17
+
+\bibitem[{{Annis} {et~al.}}(2014)]{annis14}
+{{Annis}, J.,  {Soares-Santos}, M.,  {Strauss}, M.~A.,  
+  {Becker}, A.~C.,  {Dodelson}, S.,  {Fan}, X.,  {Gunn}, J.~E.,  
+  {Hao}, J.,  {Ivezi{\'c}}, {\v Z}.,  {Jester}, S.,  {Jiang}, L.,  
+  {Johnston}, D.~E.,  {Kubo}, J.~M.,  {Lampeitl}, H.,  {Lin}, H.,  
+  {Lupton}, R.~H.,  {Miknaitis}, G.,  {Seo}, H.-J.,  {Simet}, M., \& 
+  {Yanny}, B.} 2014, \apj, 794, 120
+
+\bibitem[{{Bovy}, {Hogg} \& {Roweis}}(2009)]{bovy09}
+{Bovy}, J., {Hogg}, D.~W., \& {Roweis}, S.~T. 2011, \apj, 700, 1794
+
+\bibitem[{{Bovy}, {Hogg} \& {Roweis}}(2011)]{bovy11}
+{Bovy}, J., {Hogg}, D.~W., \& {Roweis}, S.~T. 2011, Annals of Applied Statistics, 5, 1657
+
+\end{thebibliography}
 \end{document}
